Analysis Script

### 2025-04-18 code/analysis-code/statisitical-analysis.qmd
Work

```{r}
library(ggplot2) 
library(broom) 
library(here) 
library(glmnet)
library(MASS)
library(tidymodels)
library(dplyr)
library(rsample)
library(poissonreg)
library(parsnip)
library(future)
```

@@

# NOTES TO SELF
# - change to QMD file (personally prefer)
# - work on other analysis (lag-time & quarantine fatigue? more research)
# - work on inclusion of county health ranking data or possibly not include at all? lots of mismatch counties between datasets
class(data$new_cases)
```

```{r}
# Intial Recipe setting
recipe <- recipe(new_cases ~ pop_density + transit_stations_percent_change_from_baseline + 
                     parks_percent_change_from_baseline + 
                     retail_and_recreation_percent_change_from_baseline + 
                     grocery_and_pharmacy_percent_change_from_baseline +
                     workplaces_percent_change_from_baseline + 
                     residential_percent_change_from_baseline +
                     health_factor_rank, data = data)

@@

First, I created a linear regression model as the baseline standard. This model performs very poorly as it has a low R-squared, indicating low predictive power. The model does mark all variables of the model to be significant.Additionally, the AIC is very extremely high.

@@

lm_wf <- workflow() %>%
  add_recipe(recipe)

lm_fit <- lm_wf %>%
  add_model(lm_spec) %>%
  fit(data = data)

lm_fit %>%
  extract_fit_parsnip() %>%
  tidy()

lm_fit %>%
  extract_fit_parsnip() %>%
  broom::glance()

lm_preds<- predict(lm_fit, new_data = data)

lm_metrics <- tibble(truth = data$new_cases, predicted = lm_preds$.pred) %>%
  metrics(truth = truth, estimate = predicted)
lm_metrics
```

I utilized Lasso here to hopefully bring down the RMSE and bring up the rsquared. It seems Lasso has no effect on model performance as the data lacks multicolinearity and does not have irrelevant predictors. 
```{r}
# Set method for LASSO
lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
  set_engine("glmnet")

# Insert recipe
lasso_wf <- workflow() %>%
  add_recipe(recipe)

lasso_fit <- lasso_wf %>%
  add_model(lasso_spec) %>%
  fit(data = data)

lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy()

lasso_fit %>%
  extract_fit_parsnip() %>%
  broom::glance()

lasso_preds<- predict(lasso_fit, new_data = data)

lasso_metrics <- tibble(truth = data$new_cases, predicted = lasso_preds$.pred) %>%
  metrics(truth = truth, estimate = predicted)
lasso_metrics
```

After the last two models performed poorly, the use of random forest seemed rational as this model can capture more complex relationships. 
```{r}

@@

forest_spec <- rand_forest(mode = "regression") %>%
  set_engine("ranger", seed = rngseed)

@@

# Create model with all predictors and random forest
forest_fit <- forest_wf %>%
 add_model(forest_spec) %>%
  fit(data = data)

forest_preds <- predict(forest_fit, new_data = data)
forest_preds <- tibble(truth = data$new_cases, predicted = forest_preds$.pred)
forest_rmse <- forest_preds %>%
  metrics(truth = truth, estimate = predicted)
print(forest_rmse)
```

```{r}
# Tuning forest
forest_spec <- rand_forest(
  mode = "regression", 
  trees = 300, 
  mtry = tune(), 
  min_n = tune()
) %>%
  set_engine("ranger")

# Create grid with set parameters
forest_grid <- grid_regular(
  mtry(range = c(1, 7)),  
  min_n(range = c(1, 21)),  
  levels = 7  
)

# Set seed
set.seed(1234)

# 5 fold crossvalidation repeated 5 times
# Note: Limitation of computing power so doing rolling window would not work
cv_folds <- vfold_cv(data, v = 5, repeats = 5)
plan(multisession, workers = 4)

# Tune the forest
# This step took 20-ish minutes
forest_tune_res <- tune_grid(
  forest_wf %>% add_model(forest_spec),
  resamples = cv_folds,  
  grid = forest_grid,
  metrics = metric_set(rmse, rsq),
  control = control_grid(parallel_over = "everything")
)

# Plot results
autoplot(forest_tune_res)

forest_tune_res %>%
  collect_metrics()

best_params <- select_best(forest_tune_res, metric = "rmse")
best_params

best_params$trees <- 500
```

```{r}
forest_spec <- rand_forest(
  mode = "regression",
  mtry = best_params$mtry,
  trees = best_params$trees,
  min_n = best_params$min_n
) %>%
  set_engine("ranger", seed = 1234)

forest_fit_final <- forest_wf %>%
  add_model(forest_spec) %>%
  fit(data = data)

# Get final predictions
forest_preds_final <- predict(forest_fit_final, new_data = data)
forest_preds_final <- tibble(truth = data$new_cases, predicted = forest_preds_final$.pred)

# Calculate performance metrics
forest_rmse <- forest_preds_final %>%
  metrics(truth = truth, estimate = predicted)

print(forest_rmse)
```

>>>>

Note to self:
Tried lagging mobility (doesnt improve models)
Tried filtering data to pre-vaccination (doesnt improve models)

```{r}
library(ggplot2) 
library(broom) 
library(here) 
library(glmnet)
library(MASS)
library(tidymodels)
library(dplyr)
library(rsample)
library(tibble)
library(parsnip)
library(future)
library(vip)
```

@@

@@

```{r}
# Creation of lagged cases; commonly used in epi modeling
# Note: lagged mobility does not significantly improve results from what I can tell
data <- data %>%
  group_by(county) %>%
  arrange(date) %>%
  mutate(
    lag_1 = lag(new_cases, 1),
    lag_7 = lag(new_cases, 7),
    lag_14 = lag(new_cases, 14),
  ) %>%
  ungroup()

data <- data %>% drop_na()
```

Ok, With help from ChatGPT, I implemented time blocked CV folds. The idea is that the data needs to respect the time-series nature and rolling window cannot be implemented because of the missing days in some counties.

This code splits the data into 8 folds. For each fold, the model trains on all the data up to and incuding the last fold before the test set. For example, the model will train on fold 1 and test on fold 2. Then, the model will train on folds 1 + 2 and test on fold 3.
```{r}
# Set seed
seed <- 1234
set.seed(seed)

library(dplyr)
library(lubridate)
library(rsample)

# Step 1: Ensure the data is ordered by date
train_data <- data %>%
  arrange(date)

# Step 2: Create a new variable that indicates the month or quarter
train_data <- train_data %>%
  mutate(month = floor_date(date, "month"))

# Step 3: Define the fold boundaries manually based on your specific date range
train_data <- train_data %>%
  mutate(fold_id = case_when(
    month >= as.Date("2020-03-01") & month <= as.Date("2020-06-13") ~ 1,
    month >= as.Date("2020-06-14") & month <= as.Date("2020-09-13") ~ 2,
    month >= as.Date("2020-09-14") & month <= as.Date("2020-12-13") ~ 3,
    month >= as.Date("2020-12-14") & month <= as.Date("2021-03-14") ~ 4,
    month >= as.Date("2021-03-15") & month <= as.Date("2021-06-14") ~ 5,
    month >= as.Date("2021-06-15") & month <= as.Date("2021-09-14") ~ 6,
    month >= as.Date("2021-09-15") & month <= as.Date("2021-12-14") ~ 7,
    month >= as.Date("2021-12-15") & month <= as.Date("2022-03-14") ~ 8
  ))

# Step 4: Check the number of observations per fold and the date ranges
folds_summary <- train_data %>%
  group_by(fold_id) %>%
  summarise(
    min_date = min(date),
    max_date = max(date),
    n_obs = n()
  ) %>%
  print()

# Step 5: Create the blocked CV folds using the fold_id
folds <- group_vfold_cv(train_data, group = "fold_id")

# Set aside last fold for testing
test_data <- train_data %>% filter(fold_id == 8)
train_data_final <- train_data %>% filter(fold_id < 8)

```{r}
library(rsample)
library(dplyr)
library(lubridate)
library(rsample)

# Step 0: Make sure your data is sorted
data <- data %>%
  arrange(date, county)

# Step 1: Define test set as final 30 days
test_days <- 30
last_date <- max(data$date)
test_start_date <- last_date - days(test_days - 1)

train_data <- data %>% filter(date < test_start_date)
test_data  <- data %>% filter(date >= test_start_date)

rolling_folds <- rolling_origin(
  data = train_data,
  initial = 120 * 44,  # ~4 months of training
  assess  = 30 * 44,   # ~1 month of validation
  skip    = 15 * 44,   # roll forward 1 month
  cumulative = FALSE
)

# Check date ranges for each fold
train_ranges <- purrr::map(rolling_folds$splits, ~range(pull(training(.x), date)))

# Print the ranges for each fold
train_ranges

# Step 3: (Optional) View fold summary
fold_sizes <- purrr::map_dfr(rolling_folds$splits, function(split) {
  tibble(
    train_n = nrow(training(split)),
    test_n  = nrow(testing(split))
  )
})

print(fold_sizes)

# Check date ranges too (optional)
range(train_data$date)
range(test_data$date)

```{r}
# Intial Recipe setting
recipe <- recipe(new_cases ~ pop_density +
                  retail_and_recreation_percent_change_from_baseline +
                   grocery_and_pharmacy_percent_change_from_baseline + 
                   workplaces_percent_change_from_baseline +
                   residential_percent_change_from_baseline +
                     lag_1 + lag_7 + lag_14, data = data)

recipe_baseline <- recipe(new_cases ~ pop_density + lag_1 + lag_7 + lag_14, data = data)
```

```{r}
# Define your model specification (linear regression)
base_lm_spec <- linear_reg() %>%
  set_engine("lm")

# Create the workflow with the baseline recipe
base_lm_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(base_lm_spec)

# Resampling with cross-validation
base_lm_res <- fit_resamples(
  base_lm_wf,
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

# Collect and print metrics from resampling
base_lm_metrics <- collect_metrics(base_lm_res)
print(base_lm_metrics)

# Collect and evaluate predictions
base_lm_res %>%
  collect_predictions() %>%
  group_by(id) %>%
  metrics(truth = new_cases, estimate = .pred)

# Store metrics
print(base_lm_metrics)

# Fitting the final model on the training data
base_lm_fit <- base_lm_wf %>%
  fit(data = train_data)

# Plot variable importance
vip(base_lm_fit$fit$fit)
```

@@

lm_wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(lm_spec)

# Resampling with CV
lm_res <- fit_resamples(
  lm_wf,
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

# Collect and print metrics from resampling
collect_metrics(lm_res)

lm_res %>% 
  collect_predictions() %>%
  group_by(id) %>%
  metrics(truth = new_cases, estimate = .pred)

# Store metrics
lm_metrics <- collect_metrics(lm_res)
print(lm_metrics)

# Fitting final model
lm_fit <- lm_wf %>%
  fit(data = train_data)

# Plot variable importance
vip(lm_fit$fit$fit)

After the linear regression model performed poorly, the use of random forest seemed rational as this model can capture more complex relationships and handle the quirks of the data. With an r-squared of 0.7824, this model does capture more variance than the linear regression model. Additionally, the MAE has improvements as well. The use of RMSE may be difficult due to the presence of outliers, or in other words, days where huge outbreaks occured.
```{r}
rngseed = 1234

# base_1: Forest specification (Random Forest)
base_forest_spec <- rand_forest(mode = "regression") %>%
  set_engine("ranger", seed = rngseed, importance = "impurity")

# base_2: Insert baseline recipe into workflow
base_forest_wf <- workflow() %>%
  add_recipe(recipe_baseline)

# base_3: Time-blocked CV (folds already created earlier)
# base_folds <- group_vfold_cv(train_data, group = fold_id)

# base_4: Perform cross-validation with resampling
base_forest_res <- fit_resamples(
  base_forest_wf %>% add_model(base_forest_spec),
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

# base_5: View fold performance
base_forest_res %>%
  collect_metrics() %>%
  print()

# base_6: Per-fold metrics
base_forest_res %>% 
  collect_predictions() %>%
  group_by(id) %>%
  metrics(truth = new_cases, estimate = .pred)

# base_7: Store metrics
base_forest_metrics <- collect_metrics(base_forest_res)

# base_8: Residuals
base_forest_preds <- base_forest_res %>%
  collect_predictions() %>%
  mutate(residual = new_cases - .pred)

# Plot residuals
ggplot(base_forest_preds, aes(x = residual)) +
  geom_histogram(bins = 50, fill = "#2c7fb8", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Histogram of Residuals (Random Forest)",
    x = "Residual",
    y = "Count"
  ) +
  theme_minimal()

# base_9: Fit final model on full train_data
base_forest_fit <- base_forest_wf %>%
  add_model(base_forest_spec) %>%
  fit(data = train_data)

# base_10: Variable importance plot
vip(base_forest_fit$fit$fit)
```

```{r}

@@

# Forest specification (Random Forest)
forest_spec <- rand_forest(mode = "regression") %>%
  set_engine("ranger", seed = rngseed, importance = "impurity")

@@

# Step 1: Resampling with time-blocked CV (folds from previous code)
# folds <- group_vfold_cv(train_data, group = fold_id)

# Step 2: Perform cross-validation with resampling
forest_res <- fit_resamples(
  forest_wf %>% add_model(forest_spec),    # Add forest model to workflow
  resamples = rolling_folds,                       # Use the time-blocked folds
  metrics = metric_set(rmse, rsq, mae),    # Choose evaluation metrics
  control = control_resamples(save_pred = TRUE)   # Save predictions for each fold
)

# Step 3: View fold performance
forest_res %>%
  collect_metrics() %>%
  print()

forest_res %>% 
  collect_predictions() %>%
  group_by(id) %>%
  metrics(truth = new_cases, estimate = .pred)

# Store metrics
forest_metrics <- collect_metrics(forest_res)

forest_preds <- forest_res %>%
  collect_predictions() %>%
  mutate(residual = new_cases - .pred)

ggplot(forest_preds, aes(x = residual)) +
  geom_histogram(bins = 50, fill = "#2c7fb8", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Histogram of Residuals (XGBoost)",
    x = "Residual",
    y = "Count"
  ) +
  theme_minimal()

forest_fit <- forest_wf %>%
  add_model(forest_spec) %>%
  fit(data = train_data)

vip(forest_fit$fit$fit)
```

Next, I explored the use of an XGBoost model as the Random Forest went well. These models tend to perform better. With slightly improved r-squared and similar MAE, both models are promising and are tuned in code chunks ahead.
```{r}
# base_1: Define the XGBoost model specification
base_boost_spec <- boost_tree(
  trees = 1000,
  tree_depth = 6,
  learn_rate = 0.01,
  loss_reduction = 0.01,
  sample_size = 0.8,
  mtry = 0.8
) %>%
  set_engine("xgboost", counts = FALSE) %>%
  set_mode("regression")

# base_2: Build the workflow with baseline recipe and XGBoost model
base_boost_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(base_boost_spec)

# base_3: Fit and evaluate the model using cross-validation
base_boost_res <- fit_resamples(
  base_boost_wf,
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

# base_4: View fold performance
base_boost_res %>%
  collect_metrics() %>%
  print()

# base_5: Fold-wise prediction metrics
base_boost_res %>% 
  collect_predictions() %>%
  group_by(id) %>%
  metrics(truth = new_cases, estimate = .pred)

# base_6: Collect metrics for storage
base_boost_metrics <- collect_metrics(base_boost_res)

# base_7: Collect predictions and compute residuals
base_boost_preds <- base_boost_res %>%
  collect_predictions() %>%
  mutate(residual = new_cases - .pred)

base_boost_fit <- base_boost_wf %>%
  fit(data = train_data)

vip(base_boost_fit$fit$fit)
```

```{r}
# Define the XGBoost model specification
boost_spec <- boost_tree(
  trees = 1000,
  tree_depth = 6,
  learn_rate = 0.01,
  loss_reduction = 0.01,
  sample_size = 0.8,
  mtry = 0.8
) %>%
  set_engine("xgboost", counts = FALSE) %>%
  set_mode("regression")

# Build the workflow with recipe and model
boost_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(boost_spec)

# Fit and evaluate the model using cross-validation
boost_res <- fit_resamples(
  boost_wf,
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

# Step 3: View fold performance
boost_res %>%
  collect_metrics() %>%
  print()

boost_res %>% 
  collect_predictions() %>%
  group_by(id) %>%
  metrics(truth = new_cases, estimate = .pred)

boost_metrics <- collect_metrics(boost_res)

boost_fit <- boost_wf %>%
  fit(data = train_data)

vip(boost_fit$fit$fit)
```

```{r}
# Define the Random Forest model with tuning parameters (baseline version)
base_forest_spec <- rand_forest(
  mode = "regression", 
  mtry = tune(),           
  min_n = tune(),          
  trees = 500              
) %>%
  set_engine("ranger", seed = rngseed)

# Insert baseline recipe
base_forest_wf <- workflow() %>%
  add_recipe(recipe_baseline)

# Create grid of hyperparameters for tuning
base_forest_grid <- grid_regular(
  mtry(range = c(1, 4)),  
  min_n(range = c(1, 50)), 
  levels = 10
)

# Set seed for reproducibility
set.seed(1234)

# Tune the Random Forest model using time-blocked CV (folds)
base_forest_tune_res <- tune_grid(
  base_forest_wf %>% add_model(base_forest_spec),
  resamples = rolling_folds,                     
  grid = base_forest_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(parallel_over = "everything")
)

# Plot tuning results
autoplot(base_forest_tune_res)

# Collect and print the performance metrics from tuning
base_forest_tune_res %>%
  collect_metrics() %>%
  print()

# Get best hyperparameters based on R-squared
base_best_params <- select_best(base_forest_tune_res, metric = "rsq")
print(base_best_params)

# Final baseline RF model with best parameters
base_final_forest_spec <- rand_forest(
  mode = "regression",
  mtry = base_best_params$mtry,
  min_n = base_best_params$min_n,
  trees = 500
) %>%
  set_engine("ranger", seed = rngseed)

# Final baseline workflow
base_final_forest_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(base_final_forest_spec)

# Fit final baseline model on training data
base_final_forest_fit <- base_final_forest_wf %>%
  fit(data = train_data)

base_final_forest_res <- fit_resamples(
  base_final_forest_wf,
  resamples = rolling_folds,
  metrics = metric_set(rmse, rsq, mae),
  control = control_resamples(save_pred = TRUE)
)

print(base_final_forest_res)

#test_data <- train_data %>% filter(fold_id == 8)

# Predict on test data
base_final_forest_preds <- predict(base_final_forest_fit, new_data = test_data)

# Prepare prediction results
base_final_forest_preds <- tibble(
  truth = test_data$new_cases, 
  predicted = base_final_forest_preds$.pred
)

# Evaluate performance
base_final_forest_metrics <- base_final_forest_preds %>%
  metrics(truth = truth, estimate = predicted)

print(base_final_forest_metrics)
```

```{r}
# Define the Random Forest model with tuning parameters
forest_spec <- rand_forest(
  mode = "regression", 
  mtry = tune(),           
  min_n = tune(),          
  trees = 500              
) %>%
  set_engine("ranger", seed = rngseed)

# Create grid of hyperparameters for tuning
forest_grid <- grid_regular(
  mtry(range = c(1, 8)),  
  min_n(range = c(1, 50)), 
  levels = 10
)

# Set seed for reproducibility
set.seed(1234)

# Tune the Random Forest model using time-blocked CV (folds)
forest_tune_res <- tune_grid(
  forest_wf %>% add_model(forest_spec),
  resamples = rolling_folds,                     
  grid = forest_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(parallel_over = "everything")
)

# Plot tuning results to visualize performance across hyperparameters
autoplot(forest_tune_res)

# Collect and print the performance metrics from the tuning process
forest_tune_res %>%
  collect_metrics() %>%
  print()

# Get the best hyperparameters based on the chosen metric (e.g., R-squared)
best_params <- select_best(forest_tune_res, metric = "rsq")
print(best_params)

# Train the final model with the best parameters on the full training set
final_forest_spec <- rand_forest(
  mode = "regression",
  mtry = best_params$mtry,
  min_n = best_params$min_n,
  trees = 500
) %>%
  set_engine("ranger", seed = rngseed)

# Final model workflow with the best hyperparameters
final_forest_wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(final_forest_spec)

# Fit the final model on the entire training data
final_forest_fit <- final_forest_wf %>%
  fit(data = train_data)

# Make predictions on the test data
final_forest_preds <- predict(final_forest_fit, new_data = test_data)

# Prepare prediction results
final_forest_preds <- tibble(truth = test_data$new_cases, predicted = final_forest_preds$.pred)

# Calculate RMSE for the final model
final_forest_metrics <- final_forest_preds %>%
  metrics(truth = truth, estimate = predicted)

# Print final performance metrics
print(final_forest_metrics)

final_forest_preds <- final_forest_preds %>%
  mutate(residual = truth - predicted)

ggplot(final_forest_preds, aes(x = residual)) +
  geom_histogram(bins = 50, fill = "#2c7fb8", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Histogram of Residuals (XGBoost)",
    x = "Residual",
    y = "Count"
  ) +
  theme_minimal()

```{r}
# Define the XGBoost model specification (baseline version)
base_boost_spec <- boost_tree(
  trees = 500,             # Trees to tune
  tree_depth = tune(),     # Depth of trees to tune
  learn_rate = tune(),
  loss_reduction = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Build the workflow with the baseline recipe and model
base_boost_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(base_boost_spec)

# Tune the XGBoost model using time-blocked CV (folds)
base_boost_tune_res <- tune_grid(
  base_boost_wf,
  resamples = rolling_folds,
  grid = 30,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(parallel_over = "everything")
)

# View tuning results
base_boost_tune_res %>%
  collect_metrics() %>%
  print()

# Get the best hyperparameters based on R-squared
base_best_params <- select_best(base_boost_tune_res, metric = "rsq")
print(base_best_params)

# Collect the metrics of the best model based on R-squared
base_best_metrics <- base_boost_tune_res %>%
  collect_metrics() %>%
  filter(.config == base_best_params$.config)

# Print best model metrics
print(base_best_metrics)

# Plot tuning performance
autoplot(base_boost_tune_res)

# Train the final baseline XGBoost model with best parameters
base_final_boost_spec <- boost_tree(
  trees = 500,
  tree_depth = base_best_params$tree_depth,
  learn_rate = base_best_params$learn_rate,
  loss_reduction = base_best_params$loss_reduction
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Final model workflow using the baseline recipe
base_final_boost_wf <- workflow() %>%
  add_recipe(recipe_baseline) %>%
  add_model(base_final_boost_spec)

# Fit the final XGBoost baseline model
base_final_boost_fit <- base_final_boost_wf %>%
  fit(data = train_data)

# Predict on the test data
base_final_boost_preds <- predict(base_final_boost_fit, new_data = test_data)

# Prepare results
base_final_boost_preds <- tibble(
  truth = test_data$new_cases, 
  predicted = base_final_boost_preds$.pred
)

# Calculate final performance metrics
base_final_boost_metrics <- base_final_boost_preds %>%
  metrics(truth = truth, estimate = predicted)

# Print performance metrics
print(base_final_boost_metrics)
```

```{r}
# Define the XGBoost model specification
boost_spec <- boost_tree(
  trees = 500,             # Trees to tune
  tree_depth = tune(),     # Depth of trees to tune
  learn_rate = tune(),
  loss_reduction = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Build the workflow with recipe and model
boost_wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(boost_spec)

# Tune the XGBoost model using time-blocked CV (folds)
boost_tune_res <- tune_grid(
  boost_wf,
  resamples = rolling_folds,
  grid = 30,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(parallel_over = "everything")
)

# View tuning results
boost_tune_res %>%
  collect_metrics() %>%
  print()

# Get the best hyperparameters based on R-squared
best_params <- select_best(boost_tune_res, metric = "rsq")
print(best_params)

# Collect the metrics of the best model based on R-squared
best_metrics <- boost_tune_res %>%
  collect_metrics() %>%
  filter(.config == best_params$.config)

# Print the metrics (R-squared, RMSE, MAE) for the best model
print(best_metrics)

# Plot the results of the tuning process
autoplot(boost_tune_res)

# Train the final model with the best parameters on the full training set
final_boost_spec <- boost_tree(
  trees = 500,
  tree_depth = best_params$tree_depth,
  learn_rate = best_params$learn_rate
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Final model workflow with the best hyperparameters
final_boost_wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(final_boost_spec)

# Fit the final XGBoost model on the entire training data
final_boost_fit <- final_boost_wf %>%
  fit(data = train_data)

# Make predictions on the test data
final_boost_preds <- predict(final_boost_fit, new_data = test_data)

# Prepare prediction results (truth and predicted values)
final_boost_preds <- tibble(truth = test_data$new_cases, predicted = final_boost_preds$.pred)

# Calculate RMSE, MAE, and R-squared for the final model
final_boost_metrics <- final_boost_preds %>%
  metrics(truth = truth, estimate = predicted)

# Print final performance metrics (RMSE, MAE, and R-squared)
print(final_boost_metrics)

final_boost_preds <- final_boost_preds %>%
  mutate(residual = truth - predicted)

ggplot(final_boost_preds, aes(x = residual)) +
  geom_histogram(bins = 50, fill = "#2c7fb8", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Histogram of Residuals (XGBoost)",
    x = "Residual",
    y = "Count"
  ) +
  theme_minimal()

```{r}
base_lm_metrics <- base_lm_metrics %>% 
  mutate(model = "Linear Regression (Baseline)") %>%
  select(-n, -std_err, -.config, -.estimator)

lm_metrics <- lm_metrics %>% 
  mutate(model = "Linear Regression") %>%
  select(-n, -std_err, -.config, -.estimator)

base_forest_metrics <- base_forest_metrics %>% 
  mutate(model = "Random Forest (Baseline)") %>%
  select(-n, -std_err, -.config, -.estimator)

forest_metrics <- forest_metrics %>% 
  mutate(model = "Random Forest") %>%
  select(-n, -std_err, -.config, -.estimator)

base_boost_metrics <- base_boost_metrics %>% 
  mutate(model = "XGBoost (Baseline)") %>%
  select(-n, -std_err, -.config, -.estimator)

boost_metrics <- boost_metrics %>% 
  mutate(model = "XGBoost") %>%
  select(-n, -std_err, -.config, -.estimator)

final_boost_metrics <- final_boost_metrics %>%
  mutate(model = "Tuned XGBoost") %>%
  rename(mean = .estimate)

final_forest_metrics <- final_forest_metrics %>%
  mutate(model = "Tuned Random Forest") %>%
  rename(mean = .estimate)

base_final_boost_metrics <- base_final_boost_metrics %>%
  mutate(model = "Tuned XGBoost (Baseline)") %>%
  rename(mean = .estimate)

base_final_forest_metrics <- base_final_forest_metrics %>%
  mutate(model = "Tuned Random Forest (Baseline)") %>%
  rename(mean = .estimate)

model_metrics <- bind_rows(base_lm_metrics, lm_metrics, base_forest_metrics, forest_metrics, base_boost_metrics, boost_metrics, base_final_boost_metrics, final_boost_metrics, base_final_forest_metrics, final_forest_metrics)

model_metrics <- model_metrics %>%
  select(-.estimator) %>%
  pivot_wider(names_from = '.metric', values_from = mean)
```

```{r}
library(gt)

model_table <- model_metrics %>%
  gt() %>%
  tab_header(
    title = "Model Performance Metrics",
    subtitle = "Best values are highlighted per metric"
  ) %>%
  fmt_number(columns = c(rmse, mae), decimals = 1) %>%
  fmt_number(columns = rsq, decimals = 3) %>%
  cols_label(
    model = "Model",
    rmse = "RMSE",
    rsq = "R²",
    mae = "MAE"
  ) %>%

  # Highlight lowest RMSE
  tab_style(
    style = cell_fill(color = "lightgreen"),
    locations = cells_body(
      columns = rmse,
      rows = rmse == min(rmse, na.rm = TRUE)
    )
  ) %>%

  # Highlight lowest MAE
  tab_style(
    style = cell_fill(color = "lightgreen"),
    locations = cells_body(
      columns = mae,
      rows = mae == min(mae, na.rm = TRUE)
    )
  ) %>%

  # Highlight highest R²
  tab_style(
    style = cell_fill(color = "lightgreen"),
    locations = cells_body(
      columns = rsq,
      rows = rsq == max(rsq, na.rm = TRUE)
    )
  ) %>%

  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_options(
    table.width = pct(100),
    table.align = "left",
    column_labels.padding = px(10),
    data_row.padding = px(6)
  )

library(gtExtras)
library(webshot2)

print(model_table)

gtsave(model_table, here("results", "figures", "models.png"))
```
